---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datacommons

<!-- badges: start -->
<!-- [![CRAN status](https://www.r-pkg.org/badges/version/uisapi)](https://cran.r-project.org/package=datacommons)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/uisapi)](https://cran.r-project.org/package=datacommons) -->
![R CMD Check](https://github.com/tidy-intelligence/r-datacommons/actions/workflows/R-CMD-check.yaml/badge.svg)
![Lint](https://github.com/tidy-intelligence/r-datacommons/actions/workflows/lint.yaml/badge.svg)
[![Codecov test coverage](https://codecov.io/gh/tidy-intelligence/r-datacommons/graph/badge.svg)](https://app.codecov.io/gh/tidy-intelligence/r-datacommons)
<!-- badges: end -->

Retrieve data from the [Data Commons REST API V2](https://docs.datacommons.org/api/rest/v2/). Data Commons provides harmonized access to dozens of data sources and thousands of datasets organized in a knowledge graph. 

## Installation

<!-- You can install `datacommons` from [CRAN](https://cran.r-project.org/package=datacommons) via:

```{r eval=FALSE}
install.packages("datacommons")
``` 
-->

You can install the development version of `datacommons` from GitHub with:

```{r eval=FALSE}
# install.packages("pak")
pak::pak("tidy-intelligence/r-datacommons")
```

## Usage

Load the package:

```{r}
library(datacommons)
```

Get all property labels for a given node:

```{r}
result <- dc_get_node(
    nodes = "country/USA", 
    expression = "<-"
)
str(result)
```

Find the Data Commons ID (DCID) of a place by another known ID 

```{r}
result <- dc_get_resolve(
    nodes = "Q30",
    expression = "<-wikidataId->dcid"
)
str(result)
```
